---
title: "Assignment 6: Data Scraping"
author: "Ruixin Yang"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 
library(tidyverse)
library(rvest)
library(here)
library(lubridate)

here()

# Use the theme from the lecture
mytheme <- theme_classic() +
  theme(
    axis.text = element_text(color = "black"),
    legend.position = "top"
  )

theme_set(mytheme)

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2024 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2
the_website <- read_html(
  "https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024"
)

the_website

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 
the_system_name <- the_website %>%
  html_nodes(".system-header") %>%
  html_text()

the_PWSID <- the_website %>%
  html_nodes(
    ".system-header ~ table:nth-of-type(1) tr:nth-child(1) td:nth-child(5)"
  ) %>%
  html_text()

the_ownership <- the_website %>%
  html_nodes(
    ".system-header ~ table:nth-of-type(1) tr:nth-child(2) td:nth-child(4)"
  ) %>%
  html_text()

max_withdrawals <- the_website %>%
  html_nodes(
    ".system-header ~ table:nth-of-type(10) td:nth-child(3),
     .system-header ~ table:nth-of-type(10) td:nth-child(6),
     .system-header ~ table:nth-of-type(10) td:nth-child(9)"
  ) %>%
  html_text()

the_system_name
the_PWSID
the_ownership
max_withdrawals


```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... 

5. Create a line plot of the maximum daily withdrawals across the months for 2024, making sure, the months are presented in proper sequence.

```{r create.a.dataframe.from.scraped.data}
#4 
df_withdrawals <- data.frame(
  Water_System = rep(the_system_name, 12),
  PWSID = rep(the_PWSID, 12),
  Ownership = rep(the_ownership, 12),
  Month = c(1, 5, 9, 2, 6, 10, 3, 7, 11, 4, 8, 12),
  Year = rep(2024, 12),
  Max_Withdrawals_mgd = as.numeric(max_withdrawals)
) %>%
  mutate(
    Date = my(paste(Month, "-", Year))
  )

df_withdrawals


#5 
ggplot(
  df_withdrawals,
  aes(x = Date, y = Max_Withdrawals_mgd)
) +
  geom_line() +
  scale_x_date(
    "Month",
    breaks = "1 month",
    date_labels = "%b"
  ) +
  labs(
    title = paste(
      "2024 Maximum Day Water Use for",
      the_system_name
    ),
    y = "Maximum Day Use (MGD)"
  )


```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function with two inputs - "PWSID" and "year" - that:
  - Creates a URL pointing to the LWSP for that PWSID for the given year
  - Creates a website object and scrapes the data from that object (just as you did above)
  - Constructs a dataframe from the scraped data, mostly as you did above, but includes the PWSID and year provided as function inputs in the dataframe. 
  - Returns the dataframe as the function's output


```{r construct.a.scraping.function}
#6. 
scrape.it <- function(PWSID, year) {
  
  #Construct the URL
  the_url <- paste0(
    "https://www.ncwater.org/WUDC/app/LWSP/",
    "report.php?pwsid=", PWSID,
    "&year=", year
  )
  
  #Retrieve the website contents
  the_website <- read_html(the_url)
  
  #Set the element address variables
  the_system_name_tag <- ".system-header"
  
  the_ownership_tag <- paste0(
    ".system-header ~ table:nth-of-type(1) ",
    "tr:nth-child(2) td:nth-child(4)"
  )
  
  the_data_tag <- paste0(
    ".system-header ~ table:nth-of-type(10) td:nth-child(3), ",
    ".system-header ~ table:nth-of-type(10) td:nth-child(6), ",
    ".system-header ~ table:nth-of-type(10) td:nth-child(9)"
  )
  
  #Scrape the data items
  the_system_name <- the_website %>%
    html_nodes(the_system_name_tag) %>%
    html_text()
  
  the_ownership <- the_website %>%
    html_nodes(the_ownership_tag) %>%
    html_text()
  
  max_withdrawals <- the_website %>%
    html_nodes(the_data_tag) %>%
    html_text()
  
  #Convert to a dataframe
  df_withdrawals <- data.frame(
    Water_System = rep(the_system_name, 12),
    PWSID = rep(PWSID, 12),
    Ownership = rep(the_ownership, 12),
    Month = c(1, 5, 9, 2, 6, 10, 3, 7, 11, 4, 8, 12),
    Year = rep(year, 12),
    Max_Withdrawals_mgd = as.numeric(max_withdrawals)
  ) %>%
    mutate(
      Date = my(paste(Month, "-", Year))
    )
  
  Sys.sleep(1)
  
  #Return the dataframe
  return(df_withdrawals)
}



```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2020. Then show the structure of the dataframe with either the `str()` or the `glimpse()` function. 

```{r fetch.and.plot.Durham.2020.data}
#7 
Durham.2020 <- scrape.it(
  PWSID = "03-32-010",
  year = 2020
)

glimpse(Durham.2020)

ggplot(
  Durham.2020,
  aes(x = Date, y = Max_Withdrawals_mgd)
) +
  geom_line() +
  scale_x_date(
    "Month",
    breaks = "1 month",
    date_labels = "%b"
  ) +
  labs(
    title = paste(
      "2020 Maximum Day Water Use for",
      unique(Durham.2020$Water_System)
    ),
    y = "Maximum Day Use (MGD)"
  )


```

8. Use the function above to extract data for Cary (PWSID = '03-32-010') in 2020. Combine this data with the Durham data collected above and create a plot that compares Cary's to Durham's water withdrawals.

```{r fetch.and.plot.Cary.2020.data}
#8 
Cary.2020 <- scrape.it(
  PWSID = "03-92-020",
  year = 2020
)

Durham.Cary.2020 <- bind_rows(
  Durham.2020,
  Cary.2020
)

ggplot(
  Durham.Cary.2020,
  aes(
    x = Date,
    y = Max_Withdrawals_mgd,
    color = Water_System
  )
) +
  geom_line() +
  scale_x_date(
    "Month",
    breaks = "1 month",
    date_labels = "%b"
  ) +
  labs(
    title = "Maximum Day Water Use in Durham and Cary",
    subtitle = "Monthly values for 2020",
    y = "Maximum Day Use (MGD)",
    color = "Water System"
  )


```


9. Use the code & function you created above to plot Cary's max daily withdrawal by months for the years 2018 thru 2023.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bind_rows()` to combine the dataframes into a single one, and use that to construct your plot.

```{r fetch.and.plot.Cary.multiyear.data }
#9 
the_years <- 2018:2023

the_PWSIDs <- rep.int(
  "03-92-020",
  length(the_years)
)

Cary.multiyear <- map2(
  the_PWSIDs,
  the_years,
  scrape.it
) %>%
  bind_rows()

glimpse(Cary.multiyear)

ggplot(
  Cary.multiyear,
  aes(x = Date, y = Max_Withdrawals_mgd)
) +
  geom_line() +
  geom_smooth(
    method = "loess",
    se = FALSE
  ) +
  scale_x_date(
    "Year",
    breaks = "1 year",
    date_labels = "%Y"
  ) +
  labs(
    title = "Cary Maximum Day Water Use: 2018–2023",
    y = "Maximum Day Use (MGD)"
  )



```

Question: Just by looking at the plot (i.e. not running statistics), does Cary have a trend in water usage over time?
> Answer: I think, overall, the plot suggests a slight increase in Cary’s maximum daily water use from 2018 to 2023. However, the increase is pretty small, and the trend is nearly flat. According to the plot, the month-to-month seasonal variation is much larger than the long-term trend.
