---
title: "Assignment 5: Data Visualization"
author: "Ruixin Yang"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This assignment accompanies the Environmental Data Analytics module on data
visualization. You will create, format, combine, and interpret plots using
environmental datasets. Your goal is not only to produce correct code, but also
to make deliberate choices about themes, color, accessibility, and the message
communicated by each figure.

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
7. All figures should include clear labels and units, readable text, and color
   choices appropriate for the type of variable displayed. Do not rely on color
   alone when another cue (such as shape, linetype, or facets) would make the
   figure more accessible.
8. Complete your own code before using AI in the designated comparison activity
   in Question 7. Include the prompt and your evaluation of the AI response.

---

## Set up your session 

1. Set up your session. Load the `tidyverse`, `here`, `cowplot`, and any color
palette package(s) you choose to use. Verify your project directory. Read in the
processed NTL-LTER nutrients and chemistry/physics file for Peter and Paul Lakes
(`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`) and the processed
Niwot Ridge litter file (`NEON_NIWO_Litter_mass_trap_Processed.csv`) from the
`Processed_KEY` folder.

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 
library(tidyverse)
library(here)
library(cowplot)
library(RColorBrewer)
library(lubridate)

# Read the processed datasets
PeterPaul.chem.nutrients <- read.csv(
  here(
    "Data/Processed_KEY",
    "NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"
  )
)

Niwot.litter <- read.csv(
  here(
    "Data/Processed_KEY",
    "NEON_NIWO_Litter_mass_trap_Processed.csv"
  )
)
#2 
# Convert date variables to Date format
PeterPaul.chem.nutrients <- PeterPaul.chem.nutrients %>%
  mutate(
    sampledate = ymd(sampledate)
  )

Niwot.litter <- Niwot.litter %>%
  mutate(
    collectDate = ymd(collectDate)
  )

```

## Define your theme
3. Build a theme and set it as your default theme. Start with a complete theme,
such as `theme_minimal()` or `theme_classic()`, and then customize at least two
of the following:
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

Choose a base font size that will remain readable when the plots are knitted.
Briefly identify the two or more elements you changed and explain how the
changes improve readability.

```{r build plot theme}
#3 Create a theme
course_theme <- theme_minimal(base_size = 12) +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    axis.text = element_text(color = "gray20")
  )

theme_set(course_theme)

```

> Theme choices and rationale: I used theme_minimal() and a base font size of 12. I made the plot titles bold and placed them relative to the whole plot. I also removed the minor gridlines, moved the legend to the bottom, and made the axis text darker to make it easier to read.

## Create graphs
For Questions 4–7, create `ggplot2` graphs and adjust their aesthetics using
data-visualization best practices. Match the color palette to the variable type,
label variables and units clearly, and remove formatting that does not support
the comparison.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), using
separate aesthetics for Peter and Paul lakes. Do not rely only on color: also
use shape, linetype, facets, or another appropriate visual cue. Add a separate
line of best fit for each lake using the `lm` method. Zoom in so that the central
pattern is visible without deleting the extreme observations from the analysis
(hint: use `coord_cartesian()` rather than `xlim()` or `ylim()`). Include a
descriptive title, axis labels with units, and a meaningful legend title.

```{r plot total P vs PO4}
#4 Plot total phosphorus against phosphate
phosphorus_plot <- PeterPaul.chem.nutrients %>%
  filter(
    !is.na(po4),
    !is.na(tp_ug),
    !is.na(lakename)
  ) %>%
  ggplot(aes(
    x = po4,
    y = tp_ug,
    color = lakename,
    shape = lakename
  )) +
  geom_point(alpha = 0.5) +
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    se = TRUE
  ) +
  scale_color_brewer(palette = "Dark2") +
  coord_cartesian(
    xlim = c(0, 50),
    ylim = c(0, 100)
  ) +
  labs(
    title = "Phosphate and total phosphorus in Peter and Paul Lakes",
    x = expression("Phosphate"~(mu*g/L)),
    y = expression("Total phosphorus"~(mu*g/L)),
    color = "Lake",
    shape = "Lake"
  )

phosphorus_plot

# Check observations with the highest phosphate values
PeterPaul.chem.nutrients %>%
  filter(
    !is.na(po4),
    !is.na(tp_ug)
  ) %>%
  slice_max(
    order_by = po4,
    n = 3
  ) %>%
  select(
    lakename,
    sampledate,
    depth,
    po4,
    tp_ug
  )

# Check observations with the highest total phosphorus values
PeterPaul.chem.nutrients %>%
  filter(
    !is.na(po4),
    !is.na(tp_ug)
  ) %>%
  slice_max(
    order_by = tp_ug,
    n = 3
  ) %>%
  select(
    lakename,
    sampledate,
    depth,
    po4,
    tp_ug
  )
```

Question: What relationship do you observe between phosphate and total
phosphorus? Does the relationship appear to differ between lakes? Identify any
extreme observations and explain why zooming with `coord_cartesian()` is
preferable to removing them with scale limits.

> Answer: There is a positive relationship between phosphate and total phosphorus in both lakes. But the relationship is stronger (the line is steeper) in Peter Lake, and the Paul Lake show a weaker relationship. One extreme observation from Paul Lake has a phosphate concentration of about 373.836 ug/L, with its total phosphorus concentration is about 71.066 ug/L. Peter Lake has observations with total phosphorus around 146.782 ug/L. The `coord_cartesian()` function only clips the final displayed image and does not delete the data. The regression line still uses all the data for calculation. However, xlim() and ylim() will first remove the data outside the range, and then calculate the regression line. Therefore, it may change the regression result.


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c)
TN, with month on the x-axis and lake represented by a consistent aesthetic.
Use a qualitative, color-vision-friendly palette. Then create a `cowplot` that
combines the three graphs. Make sure that only one legend is present, the month
order is correct, graph axes are aligned, and each y-axis states the variable
and units.

Tips:
* Month should appear in calendar, not alphabetical, order. You will need to convert 
to factor to order it.
* Setting an axis title in your theme to `element_blank()` removes the axis title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.

```{r Create boxplots, fig.height=9, fig.width=6}
#5 Create an ordered month variable
PeterPaul_month <- PeterPaul.chem.nutrients %>%
  mutate(
    Month = factor(
      format(sampledate, format = "%b"),
      levels = c(
        "May", "Jun", "Jul", "Aug",
        "Sep", "Oct", "Nov"
      )
    )
  )

# Temperature boxplot
temperature_boxplot <- PeterPaul_month %>%
  filter(
    !is.na(temperature_C),
    !is.na(Month)
  ) %>%
  ggplot(aes(
    x = Month,
    y = temperature_C,
    fill = lakename
  )) +
  geom_boxplot(
    width = 0.6,
    alpha = 0.8,
    outlier.alpha = 0.35
  ) +
  scale_fill_brewer(palette = "Dark2") +
  scale_x_discrete(drop = FALSE) +
  labs(
    title = "Water temperature by month",
    x = NULL,
    y = "Temperature (degrees C)",
    fill = "Lake"
  ) +
  theme(legend.position = "none")

# Total phosphorus boxplot
tp_boxplot <- PeterPaul_month %>%
  filter(
    !is.na(tp_ug),
    !is.na(Month)
  ) %>%
  ggplot(aes(
    x = Month,
    y = tp_ug,
    fill = lakename
  )) +
  geom_boxplot(
    width = 0.6,
    alpha = 0.8,
    outlier.alpha = 0.35
  ) +
  scale_fill_brewer(palette = "Dark2") +
  scale_x_discrete(drop = FALSE) +
  labs(
    title = "Total phosphorus by month",
    x = NULL,
    y = expression("Total phosphorus"~(mu*g/L)),
    fill = "Lake"
  ) +
  theme(legend.position = "none")

# Total nitrogen boxplot
tn_boxplot <- PeterPaul_month %>%
  filter(
    !is.na(tn_ug),
    !is.na(Month)
  ) %>%
  ggplot(aes(
    x = Month,
    y = tn_ug,
    fill = lakename
  )) +
  geom_boxplot(
    width = 0.6,
    alpha = 0.8,
    outlier.alpha = 0.35
  ) +
  scale_fill_brewer(palette = "Dark2") +
  scale_x_discrete(drop = FALSE) +
  labs(
    title = "Total nitrogen by month",
    x = "Month",
    y = expression("Total nitrogen"~(mu*g/L)),
    fill = "Lake"
  )

plot_grid(
  temperature_boxplot,
  tp_boxplot,
  tn_boxplot,
  nrow = 3,
  align = "h",
  rel_heights = c(1, 1, 1.2)
)

```

Question: What do you observe about the center, spread, and potential outliers
of each variable across seasons and between lakes? Support your interpretation
with specific features visible in the boxplots.

> Answer: Water temperature generally increases from May through August or September and then decreases in October and November. The two lakes have similar temperature distributions, but Paul Lake has a slightly higher median in several months. In the box plot for summer, the wide box and the whiskers indicate greater temperature fluctuations. Total phosphorus is generally higher in Peter Lake, especially from July through September. Peter Lake also has a wider spread and more high-value outliers. Paul Lake usually has a lower total nitrogen content and less fluctuation. Total nitrogen shows a similar pattern. Peter Lake generally has a higher median and greater spread, particularly during the summer. Both lakes contain potential high-value outliers, but the most extreme TN observations occur in Peter Lake.

6. [Niwot Ridge] Subset the litter dataset to the `"Needles"` functional group.
Plot dry mass of needle litter by date and distinguish NLCD classes using a
qualitative color aesthetic. You do not need to rename the land-use classes.
Format the date axis, labels, legend, point size, and transparency so that
overlapping observations remain visible.

7. [Niwot Ridge] First, independently create a second version of the plot with
the three NLCD classes separated into facets rather than distinguished by color.
Keep other formatting choices as consistent as possible so that you can compare
the effect of faceting itself.

```{r Plot litter}
#6 Keep only needle litter
Niwot_needles <- Niwot.litter %>%
  filter(
    functionalGroup == "Needles",
    !is.na(collectDate),
    !is.na(dryMass),
    !is.na(nlcdClass)
  )

needles_color_plot <- Niwot_needles %>%
  ggplot(aes(
    x = collectDate,
    y = dryMass,
    color = nlcdClass,
    shape = nlcdClass
  )) +
  geom_point(
    size = 1.4,
    alpha = 0.5
  ) +
  scale_color_brewer(palette = "Dark2") +
  scale_x_date(
    date_breaks = "4 months",
    date_labels = "%b\n%Y",
    expand = expansion(mult = c(0.01, 0.03))
  ) +
  labs(
    title = "Needle litter dry mass through time",
    x = "Collection date",
    y = "Dry mass (g)",
    color = "NLCD class",
    shape = "NLCD class"
  )

needles_color_plot


#7 Separate NLCD classes into facets
needles_facet_plot <- Niwot_needles %>%
  ggplot(aes(
    x = collectDate,
    y = dryMass
  )) +
  geom_point(
    size = 1.4,
    alpha = 0.5,
    color = "gray45"
  ) +
  facet_wrap(
    vars(nlcdClass),
    ncol = 1
  ) +
  scale_x_date(
    date_breaks = "4 months",
    date_labels = "%b\n%Y",
    expand = expansion(mult = c(0.01, 0.03))
  ) +
  labs(
    title = "Needle litter dry mass through time by NLCD class",
    x = "Collection date",
    y = "Dry mass (g)"
  ) +
  theme(legend.position = "none")

needles_facet_plot


```

## AI comparison

8. After completing your own faceted plot, provide an AI tool with the Question 7 instructions and the relevant variable names. Ask it to propose `ggplot2` code for the same figure. Do **not** replace your original solution with the AI version.

Paste the exact prompt you used below:

> AI tool used: ChatGPT

> AI prompt: Here are my files. Using ggplot2, filter the dataset to observations where functionalGroup == "Needles". Create a plot of dryMass over collectDate, with the three nlcdClass categories separated into facets rather than distinguished by color. Keep the same y-axis scale across the facets, make overlapping observations visible, and include clear date formatting, labels, units, and a descriptive title. Provide R code.

Run and inspect the AI-generated code, correcting only errors required to make
it run. If you make corrections, identify them. Then compare the two solutions.

```{r AI solution, fig.width=8, fig.height=4}
# Paste the AI-generated code here. Clearly comment any corrections you make.

# AI-generated solution
ai_facet_plot <- Niwot.litter %>%
  filter(
    functionalGroup == "Needles",
    !is.na(collectDate),
    !is.na(dryMass),
    !is.na(nlcdClass)
  ) %>%
  ggplot(
    aes(
      x = collectDate,
      y = dryMass
    )
  ) +
  geom_point(
    color = "#0072B2",
    size = 1.8,
    alpha = 0.55
  ) +
  facet_wrap(
    vars(nlcdClass),
    nrow = 1
  ) +
  scale_x_date(
    date_breaks = "1 year",
    date_labels = "%Y"
  ) +
  labs(
    title = "Needle litter dry mass across NLCD classes",
    subtitle = "Each panel represents one land-cover class",
    x = "Collection date",
    y = "Dry mass (g)"
  ) +
  theme_classic(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold"),
    strip.background = element_rect(
      fill = "grey90",
      color = "grey70"
    ),
    strip.text = element_text(face = "bold")
  )

ai_facet_plot

```

9. Which plot—your solution or the AI solution—more effectively communicates the data? Evaluate at least three specific choices, such as data filtering, facet structure, date-axis formatting, labels and units, theme, accessibility, or unnecessary code. Identify one useful idea from the AI
solution and one choice you would reject or improve. If the solutions are very similar, explain why both approaches are appropriate.

> AI comparison: I think the AI solution communicates the data more effectively overall. Both ways gave solutions to those requirements. However, the AI solution places the three facets in one horizontal row. This layout is more compact and makes it easier to compare the classes side by side. My vertical layout gives each class more horizontal space for showing changes over time. In addition, the AI solution uses blue points on a white background, which provides stronger contrast and makes the observations easier to see. In my solution, the gray points blend into the gray background and gridlines, so the graph looks less clear. In terms of code, my solution is more concise because it reuses the Niwot_needles dataset created in Q6 and relies on the default theme. The AI solution repeats the filtering step and includes several additional theme settings, so its code is longer. One useful idea from the AI solution is the combination of horizontal facets, a simple theme, and high-contrast points. However, I would like to keep or improve a little bit for my date label because it can more clearly present the time information for the entire year. Overall, I prefer the AI solution because its tidy and more easier to read for me.

10. Question: Which encoding in Questions 6 and 7—color or facets—do you think is more effective for comparing NLCD classes, and why? Consider overlap, ease of comparison, accessibility, and the amount of space used by each figure.

> Answer: I think facets are more effective than color for comparing the three NLCD classes in this dataset. In the color-and-shape version, many observations occur on similar dates and have low values, so points from different classes can overlap. This makes it harder to identify the pattern within each class. Faceting separates the observations into different panels, so it will be easier to compare the amount and range for dry mass. In addition, the color version is more compact, but due to the overlapping, it becomes more difficult to distinguish each point. Although facets require more space and require the reader to look between panels, I think the improvement in clarity is more important. Therefore, I would choose facets.
